# Mixed Effects Model {#sec-longi-mixeffect}

```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
knitr::opts_chunk$set(fig.height = 4)
knitr::opts_chunk$set(fig.width = 6)
knitr::opts_chunk$set(fig.align="center")
`%notin%` <- Negate(`%in%`)
```

```{r, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(haven))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(kableExtra))
```

```{r, echo=FALSE}
TLC <- read.csv("Data/TLC.csv")

### Convert Data from wide to Long
long_TLC <- tidyr::gather(TLC, level, measurements, lead0:lead6, factor_key=TRUE)
long_TLC <- long_TLC[order(long_TLC$id),]

#create new time variable for analyses
long_TLC$time <- c(0,1,4,6)[long_TLC$level]

#Note: other ways to code for time as continuous
# create new variable to make time continuous
long_TLC$week = 0 #baseline, week 0
long_TLC$week[long_TLC$level == "lead1"] = 1 #week 1
long_TLC$week[long_TLC$level == "lead4"] = 4 #week 4
long_TLC$week[long_TLC$level == "lead6"] = 6 #week 6
```

**Learning Objectives**

1. Understand how to fit mixed effects models to longitudinal data
2. Understand how to interpret results

## Introduction
Lets consider a few simple cases in understanding the Mixed-Effect Model:

$$\mu_{ij} = E(Y_{ij} \mid {X_{ij},t_{ij}})$$

where $X_{ij}$ are the explanatory variables and $t_{ij}$ is time. 


**Individual intercepts:** The coefficient $b_i$ represents each $i$th individual starting level.

Consider the three following cases:

a) $\mu_{ij} = \mu + b_i$

b) Model with only time: $\mu_{ij} = \beta_0 + \beta_1 t_{ij} + b_i$. The systematic component is $\beta_0 + \beta_1 t_{ij}$ and random component is $b_i$

c) Model with only covariates: $\mu_{ij} = \beta_0 + b_i + \beta_1 X_{ij}$. The systematic component is $\beta_0 + \beta_1 X_{ij}$ and random component is $b_i$

In all of these cases (a-c), the coefficient $b_i$ represents the difference between the overall mean intercept and an individual's intercept. For example, if $b_i = -2$ then the individual $i$ starts at $2$ units lower in their outcome on average.

In the Mixed-Effects model, we assume that each individual has their own intercept $b_i$, where $b_i \overset{\text{iid}}{\sim} N(0,\sigma^2_b)$ and where $\sigma^2_b$ can be obtained from the summary R output. In addition, we assume that errors have constant variance $e_{ij} \overset{\text{iid}}{\sim} N(0,\sigma^2_e)$ where $\sigma^2_e$ can be obtained from the summary R output. 

Using the TLC data as an example, consider the two models

1) a simple Mixed-Effects model with a random  intercept is defined as:

$$E(y_{ij} \mid X_{ij}) = \beta_0 +  \beta_{trt}trt +  \beta_{T}T + b_{0,i}$$

**Individual slopes:** The interpretation of a random slope in the model is that each individual has a unique growth rate or trajectory.

2) a simple Mixed-Effects model with a random  intercept and random slope for time is defined as:

$$E(y_{ij} \mid X_{ij}) = \beta_0 +  \beta_{trt}trt +  \beta_{T}T + b_{0,i} + b_{T,i}T$$

The systematic component is $\beta_0 +  \beta_{trt}trt + \beta_{T}T$ and random component is $b_{0,i} + b_{T,i}T$. Here every $i$th individual has a unique starting point $b_{0,i}$, hence the average starting point of blood lead levels (BLL) for each individual is $\beta_0  + b_{0,i}$. Further, each $i$th individual has a unique growth trajectory of BLL $b_{T,i}$ where the average changes in BLL is given by $\beta_1  + b_{T,i}$.Thus, the new slope for time is $(\beta_{T} + b_{T,i})T$

Note: If the growth rate is explained by explanatory variables ($X_{ij}$) then include an interaction term. Remember that explanatory variables in the random component of the Mixed-Effects model **must be time-varying**. 

## Brief Review of what we learned so far 

**Models for LDA with Continuous Outcomes**

* Marginal Models
  + GLS
    + Define covariance model
    + Compare models using AIC, BIC, likelihood ratio test
  + GEE
    + Compare models using QIC
    + Define covariance model and link function


* \textcolor{blue}{Mixed-Effects Models}
  + Specify random effects
  + Compare models using AIC, BIC 
  + Other packages allow you to specify the covariance model


## Section 4: Fitting Mixed Effect Models to the TLC data

```{r}
library(tidyr) #Allows for us to manipulate the data structure
library(data.table) #Allows for us to manipulate the data structure
library(lme4) #Allows us to fit mixed effects models
library(lattice) #for plotting random effects
```

Here we will learn how to use R to fit Mixed-Effects (subject-specific) Models for longitudinal data, when the outcome of interest is a continuous variable (linear models).

\begin{center}
Model 1: $E(y_{ij} \mid X_{ij}) = \beta_0 +  \beta_{trt}trt +  \beta_{T}T + b_{0,i}$
\end{center}

The (1|id) means that we are allowing the intercept, represented by 1, to vary by patient.

```{r}
#label this model object as mod_1
mod_1 <- lmer(measurements ~ 1 + group + time + (1|id), 
              data = long_TLC, REML=TRUE)
summary(mod_1)
```


```{r, eval=TRUE}
# added  - some helpful functions
coef(mod_1) # plot all coefficients for each individual
ranef(mod_1) # print random effects, in this case, random intercept
```


```{r}
# added  - some helpful functions
fixef(mod_1) # print all fixed effects
AIC(mod_1) #print model AIC
summary(ranef(mod_1)$id) # calculate summary statistics of random intercept
boxplot(ranef(mod_1)$id) # create boxplot of estimated random intercept
```


```{r}
# added  - some helpful functions
summary(resid(mod_1))  # calculate summary statistics of residuals
hist(resid(mod_1)) # create histogram to assess distribution of residuals
```


```{r}
# add plot to visualize individual intercepts using dotplot() from the lattice package
dotplot(ranef(mod_1))
```


\begin{center}
Model 2: $E(y_{ij} \mid X_{ij}) = \beta_0 +  \beta_{trt}trt +  \beta_{T}T + b_{0,i} + b_{T,i}T$
\end{center}

Here the term (1+time|id) specifies the random intercept and random slope for time.

```{r}
#label this model object as mod_2
mod_2 <- lmer(measurements ~  1 + group + time+ (1 + time|id),
              data = long_TLC, REML = TRUE)
summary(mod_2)
```

\begin{center}
Model 3: $E(y_{ij} \mid X_{ij}) = \beta_0 + \beta_{trt}trt + \beta_{T}T + \beta_{trt:T}trt*T + b_{0,i}$
\end{center}

This model includes time as a continuous covariate and also includes an interaction term to account and adjust for any possible discrepancies between the treatment and control.

```{r}
#label this model object as mod_3
mod_3 <- lmer(measurements ~ 1 + group + time + group:time + (1|id),
              data = long_TLC, REML = TRUE)
summary(mod_3)
```

\begin{center}
Model 4: $E(y_{ij} \mid X_{ij}) = \beta_0 +  \beta_{trt}trt +  \beta_{T}T + \beta_{trt:T}T*trt+b_{0,i} + b_{T,i}T$
\end{center}

This model treats time as a continuous covariate and also introduces a quadratic term.

```{r}
#label this model object as mod_4
mod_4 <- lmer(measurements ~ 1 + group + time + time:group + (1 + time|id), 
              data = long_TLC, REML = TRUE)
summary(mod_4)
```


\begin{center}
Model 5: $E(y_{ij} \mid X_{ij}) = \beta_0 + \beta_{trt}trt + \beta_{T}T + \beta_{T^2}T^2 + \beta_{trt:T}trt*T+\beta_{trt:T^2}trt*T^2+b_{0,i} + b_{T,i}T$
\end{center}

This model includes time as a continuous covariate and also includes a quadratic term to capture any possible non-linearity in time and an interaction term to account and adjust for any possible discrepancies between the treatment and control.

```{r}
#label this model object as mod_5
mod_5 <- lmer(measurements ~  1 + group + time + I(time^2) + group:(time + I(time^2)) +
                (1+time|id) + (I(time^2)|id),  data = long_TLC, REML = TRUE)
summary(mod_5)
```

```{r}
mod_5_test <- lmer(measurements ~  1 + group + time + I(time^2) + group:(time + I(time^2)) +
                   (1 + time + I(time^2)|id),  data = long_TLC, REML = TRUE)
summary(mod_5_test)
```

## Model Comparison

We are able to compare the models using AIC and BIC. After you write the other models above, run this chunk below.

```{r}
Mixed_Model_comparisons <- data.frame("Model Name" = c("Mixed Model 1", "Mixed Model 2", "Mixed Model 3",
                                                       "Mixed Model 4", "Mixed Model 5"),
                                      "AIC" = c(AIC(mod_1), AIC(mod_2), AIC(mod_3),
                                                AIC(mod_4), AIC(mod_5)),
                                      "BIC" = c(BIC(mod_1), BIC(mod_2), BIC(mod_3),
                                                BIC(mod_4), BIC(mod_5)))
Mixed_Model_comparisons
```

Which model would you select?

**Answer: We would pick Model 5 because it has the lowest AIC and lowest BIC value.**
